\vspace{-15mm}\fontsize{21pt}{10pt}\selectfont\textbf{Wordy Manuscript Title: \\Even Wordier Manuscript Subtitle\thanks{Draft manuscript. Please do not cite without the authors' permission.}}

### 2022-04-22 sections/methods.tex
Add: Supplementary section for consensus network

 % TODO: Use bold for matrices and lower-case bold for vectors
  This section defines the notation used below to describe the consensus network algorithm used in the our pipeline. Note that all networks to be compared have the same number of nodes.

  $w$ = number of co-occurrence networks to be integrated in the consensus network (corresponding to the number of the different approaches used to infer co-occurrence networks)

  $q$ = number of nodes in any co-occurrence network (assumed to be the same across all $w$ networks)

  $N^i$ = matrix of network edge Weights for the $i^{th}$ co-occurrence network. This is a $q \times q$ matrix, where $i \in \{1,\dots,w\}$. $N^i_{a,b}$ represents edge $(a,b)$ in network $i$

  $P^i$ = matrix of p-values for all edges of the $i^{th}$ co-occurrence network. This is a $q \times q$ matrix, where $i \in \{1,\dots,q\}$

@@

  % TODO: Is this necessary here, should this be elaborated?
  As described in more detail in the Supplementary Material and in the original reference~\cite{Poole_Gibbs_Shmulevich_Bernard_Knijnenburg_2016}, the final combined p-value is given by:
  \begin{equation}
    \begin{aligned}
        & \hat{P}_j = 1.0 - \Phi_{2f}\left( \psi / c \right) \\
        \text{where},~ &\psi = -2 \sum_{i=1}^k \log(\bar{P}^i_j) ~~~\text{and}~~~ \Phi_{2f} = \mathrm{CDF}\left( \chi^2_{2f} \right)
    \end{aligned}
    \label{eqn:pvalue-combined}
  \end{equation}
  where, $\hat{P}_j$ is the combined p-value for the edge $j$, f is the number of degrees of freedom, and c is a scale factor.

  % FIXME: Not sure what our stance on this is?
  Note that we do not use Pearson and Spearman methods in the p-value merging step and these algorithms are only used for demonstration and comparison.
  The combined p-values are used to threshold for significance right before the consensus algorithm is applied to the inferred networks.

@@

  \begin{equation*}
    g(x) =
    \begin{cases}
       & 0, \text{ if } x=0, \\
       & -1, \text{ if } x<0, \\
       & 1, \text{ if } x>0
    \end{cases}
  \end{equation*}
  and
  \begin{equation*}
    f(x_1,\dots,x_w) = max \left( \#(i \mid x_i=-1),\#(i \mid x_i=1)
       \right)
  \end{equation*}

@@

  The simple voting method gives the union of the networks when $p \in (0, \frac{1}{w})$ and will return the intersection when $p \in ((w - 1), 1)$.
  In general, if $p \in ((n - 1), \frac{n}{w})$, this algorithm will report an edge in the consensus network when at least $n$ network inference methods report this edge. [REFERENCE]

@@

  \begin{equation}
    s_j = \sum_i \bar{S}^i_j
    \label{eqn:scaled-sum}
  \end{equation}

@@

  \subsubsection*{Notation}
  \vspace{-5mm}
  This section defines notation for the network variability analysis in our study\dots
  % TODO: Update W and Q in the entire subsection
  $W$ =  number of co-occurrence networks generated from all possible combinations of tools and parameters in the workflow. Note that this is different from $w$, which counted only the possible network inference modules.

>>>>

  This section defines the notation used below to describe the consensus network algorithm used in the our pipeline. Note that all networks to be compared have the same number of nodes.

  $w$, the number of co-occurrence networks to be integrated in the consensus network (corresponding to the number of the different approaches used to infer co-occurrence networks)

  $q$, the number of nodes in any co-occurrence network (same across all $w$ networks)

  $N^i$, the matrix of network edge weights for the $i^{th}$ co-occurrence network. This is a $q \times q$ matrix, where $i \in \{1,\dots,w\}$. $N^i_{a,b}$ represents edge $(a,b)$ in network $i$

  $P^i$, the matrix of p-values for all edges of the $i^{th}$ co-occurrence network. This is a $q \times q$ matrix, where $i \in \{1,\dots,q\}$

@@

  As described in more detail in the Supplementary section and in the original reference~\cite{Poole_Gibbs_Shmulevich_Bernard_Knijnenburg_2016}, the final combined p-value is given by:
  \begin{equation}
    \begin{aligned}
        & \hat{P}_j = 1.0 - \Phi_{2f}\left( \psi / c \right) \\
        \text{where},~ &\psi = -2 \sum_{i=1}^w \log(\bar{P}^i_j) ~~~\text{and}~~~ \Phi_{2f} = \mathrm{CDF}\left( \chi^2_{2f} \right)
    \end{aligned}
    \label{eqn:pvalue-combined}
  \end{equation}
  where, $\hat{P}_j$ is the combined p-value for the edge $j$, f is the number of degrees of freedom, and c is a scale factor.

  Note that we do not use Pearson and Spearman methods in the p-value merging step to calculate the consensus network, these algorithms are only used for demonstration and comparison.
  The combined p-values are used to threshold for significance right before the consensus algorithm is applied to the inferred networks.

@@

  \begin{equation*}
    g(x) =
    \begin{cases}
       & 0, \text{ if } x=0, \\
       & -1, \text{ if } x<0, \\
       & 1, \text{ if } x>0
    \end{cases}
  \end{equation*}
  and
  \begin{equation*}
    f(x_1,\dots,x_w) = max \left( \#(i \mid x_i=-1),\#(i \mid x_i=1)
       \right)
  \end{equation*}
  where, $#$ refers to the cardinality of the set.

@@

  The simple voting method gives the union of the networks when $p \in (0, \frac{1}{w})$ and will return the intersection when $p \in ((w - 1), 1)$.
  In general, if $p \in ((n - 1), \frac{n}{w})$, this algorithm will report an edge in the consensus network when at least $n$ network inference methods report this edge.

@@

  \begin{equation}
    s_j = \sum_{i=1}^{w} \bar{S}^i_j
    \label{eqn:scaled-sum}
  \end{equation}

@@

  \subsubsection*{Notation}
  \vspace{-5mm}
  This section defines notation for the network variability analysis in our study\dots

  % TODO: Update W and Q in the entire subsection
  $W$ =  number of co-occurrence networks generated from all possible combinations of tools and parameters in the workflow. Note that this is different from $w$, which counted only the possible network inference modules.

### 2022-04-22 sections/results.tex
Add: Supplementary section for consensus network

  % TODO: Update after getting SparCC results
  In order to obtain a quantitative picture of the differences between the inferred networks, we analyzed the distribution of common nodes and edges (Figure~\ref{fig:figure5} B and C) using UpSet plots~\cite{Lex}.
  Only \ac{harmonies}, \ac{cozine},\ac{spieceasi}, \ac{spring}, propr, and \ac{sparcc} are used in the comparison since Pearson and Spearman add a large number of spurious edges since they are not intended for compositional datasets).
  The results for the node intersections show that the networks have a large number of nodes in common ($33$ out of $68$ total nodes) and no network possesses any unique node.
  The edge intersections in contrast show that only $8$ edges (out of $202$ total edges) are in common between all the methods and each network has a large number of unique edges.
  These results indicate that there is a substantial rewiring of connections in the inferred networks.

  % TODO: What is the final concluding statement here? How to segway to the next section

@@

  We observe a large number of rewirings in the inferred association networks generated by the different methods (Figure \ref{fig:figure5}C), implying that there is a lack of agreement on the associations that should be a part of the final network.
  Therefore, we have developed a consensus method that would take into consideration the evidence offered by each network inference algorithm and generate a consensus network that would contain the most probable interactions for the input data.
  % TODO: Should we include Pearson and Spearman in the consensus (currently we do)!
  We have developed two algorithms for the generation of the consensus network, the simple voting and the scaled-sum methods (see Methods).
  First, for the correlation-based methods we merge the p-values using the Brown's p-value merging method~\cite{Poole_Gibbs_Shmulevich_Bernard_Knijnenburg_2016,faustCoNetAppInference2016} (see Methods).
  Second, we filter all the inferred networks based on an association strength threshold of 0.1 and a p-value cutoff of 0.05.
  Finally, we apply the consensus algorithms we have developed on these filtered networks.
  The simple voting method does a consensus based on whether the number of inferred networks that have an edge between a pair of nodes is greater than a threshold.
  The scaled-sum method does a consensus based on whether the normalized sum of an edge between a pair of nodes from all inferred networks is greater than a threshold.
  The default algorithm for the pipeline is the scaled-sum method, since this method takes into account the weights of the association in the determination of the final consensus.
  Additionally, by default we include all the network inference methods in the consensus calculation by default, but a subset of methods can be considered if it is required.
  Based on this approach, \ac{micone} reports as default output the consensus network, where is each edge is annotated with weights (correlations for the correlation-based methods and direct associations for the other methods) from all the methods used in the consensus algorithm.

@@

  % TODO: Supplementary work: Figure S6
  The precision and sensitivity graphs on the seqtime data can be found in the Supplementary section.

@@

  Figure~\ref{fig:figure7}A shows the default network compared against networks generated by altering one of the steps of the pipeline from the default.
  These results indicate that the biggest rewiring in networks occur when the reference database or the network inference algorithm are changed.
  % TODO: Show proof??
  The filtering in the \ac{OP} step when turned off also adds a large number of new nodes, but the edges between the common nodes (with the default network) are fairly conserved.
  Furthermore, the L1 distance of networks generated by altering one of the steps of the pipeline from the default against the default network (Figure~\ref{fig:figure7}B) shows that the biggest deviations from the default network occur when the \ac{ni} steps are changed.
  This result is different from that observed in Figure~\ref{fig:figure2} since in this particular analyses all the steps other than the \ac{ni} step have the consensus algorithm applied to them, which drastically reduces the edges in the graph that have lower probabilities.
  This result reinforces the importance of the consensus network in increasing the robustness of the inferred networks to different processing methods.
  Figure~\ref{fig:figure7}C shows the co-occurrence networks inferred for the healthy subjects (control) and autistic subjects in the fecal microbiome transplant study~\cite{Kang2017}.
  These consensus networks were generated using the default tools and parameters from Table~\ref{tab:default_options}.

>>>>

  In order to obtain a quantitative picture of the differences between the inferred networks, we analyzed the distribution of common nodes and edges (Figure~\ref{fig:figure5} B and C) using UpSet plots~\cite{Lex}.
  Only \ac{harmonies}, \ac{cozine},\ac{spieceasi}, \ac{spring}, propr, and \ac{sparcc} are used in the comparison since Pearson and Spearman add a large number of spurious edges since they are not intended for compositional datasets).
  The results for the node intersections show that the networks have a large number of nodes in common ($33$ out of $68$ total unique nodes) and no network possesses any unique node.
  The edge intersections in contrast show that only $8$ edges (out of $202$ total unique edges) are in common between all the methods and each network has a large number of unique edges.
  These results indicate that there is a substantial rewiring of connections in the inferred networks.
  Hence, in order to determine the interactions that are most likely to exist in the dataset we advocate for the construction of a consensus algorithm.

@@

  We observe a large number of rewirings in the inferred association networks generated by the different methods (Figure \ref{fig:figure5}C), implying that there is a lack of agreement on the associations that should be a part of the final network.
  Therefore, we have developed a consensus method that would take into consideration the evidence offered by each network inference algorithm and generate a consensus network that would contain the most probable interactions for the input data.
  We have developed two algorithms for the generation of the consensus network, the simple voting and the scaled-sum methods (see Methods).
  We only include \ac{sparcc} and propr in the consensus network calculations, for consensus networks that include Spearman and Pearson see Figure~\ref{fig:figure_s6} in the Supplementary section.
  First, for the correlation-based methods we merge the p-values using the Brown's p-value merging method~\cite{Poole_Gibbs_Shmulevich_Bernard_Knijnenburg_2016,faustCoNetAppInference2016} (see Methods).
  Second, we filter all the inferred networks based on an association strength threshold of 0.1 and a p-value cutoff of 0.05.
  Finally, we apply the consensus algorithms we have developed on these filtered networks.
  The simple voting method does a consensus based on whether the number of inferred networks that have an edge between a pair of nodes is greater than a threshold.
  The scaled-sum method does a consensus based on whether the normalized sum of an edge between a pair of nodes from all inferred networks is greater than a threshold.
  The default algorithm for the pipeline is the scaled-sum method, since this method takes into account the weights of the association in the determination of the final consensus.
  Additionally, by default we include all the network inference methods in the consensus calculation by default, but a subset of methods can be considered if it is required.
  Based on this approach, \ac{micone} reports as default output the consensus network, where is each edge is annotated with weights (correlations for the correlation-based methods and direct associations for the other methods) from all the methods used in the consensus algorithm.

@@

  % TODO: Supplementary work: Figure S7
  The precision and sensitivity graphs on the seqtime data can be found in the Supplementary section.

@@

  Figure~\ref{fig:figure7}A shows the default network compared against networks generated by altering one of the steps of the pipeline from the default.
  These results indicate that the biggest rewiring in networks occur when the reference database or the network inference algorithm are changed.
  % TODO: Show proof??
  The filtering in the \ac{OP} step when turned off also adds a large number of new nodes, but the edges between the common nodes (with the default network) are fairly conserved.
  % TODO: Say that this shows how important the consensus network is in the whole pipeline
  Furthermore, the L1 distance of networks generated by altering one of the steps of the pipeline from the default against the default network (Figure~\ref{fig:figure7}B) shows that the biggest deviations from the default network occur when the \ac{ni} steps are changed.
  This result is different from that observed in Figure~\ref{fig:figure2} since in this particular analyses all the steps other than the \ac{ni} step have the consensus algorithm applied to them, which drastically reduces the edges in the graph that have lower probabilities.
  This result reinforces the importance of the consensus network in increasing the robustness of the inferred networks to different processing methods.
  Figure~\ref{fig:figure7}C shows the co-occurrence networks inferred for the healthy subjects (control) and autistic subjects in the fecal microbiome transplant study~\cite{Kang2017}.
  These consensus networks were generated using the default tools and parameters from Table~\ref{tab:default_options}.

### 2022-04-22 sections/supplementary.tex
Add: Supplementary section for consensus network

@@

@@

  % FIXME: Update according to the new notation in methods section
  \subsubsection*{p-value merging}
  Fisher~\cite{fisher_224a_1948} proposed that for $k$ independent p-values, each generated by $k$ different methods and denoted by $P_i$, the statistic $\Psi$:
  \begin{equation*}
    \begin{aligned}
        \Psi &= \sum_{i=1}^k -2 \log \left( P_i \right) \\
        \Psi &\sim \chi^2_{2k}
    \end{aligned}
  \end{equation*}

  Brown~\cite{brown_400_1975} extended Fisher's method to dependent p-values by using a re-scaled $\chi^2$ distribution:
  \begin{equation*}
    \Psi \sim c \chi^2_{2f}
  \end{equation*}
  where, $f$ is the degrees of freedom and $c$ is the scale factor and are given by:
  \begin{equation*}
    f = \frac{\mathrm{E}[\Psi]^2}{\mathrm{Var}[\Psi]} ~~~\text{and}~~~ c = \frac{\mathrm{Var}[\Psi]}{2\mathrm{E}[\Psi]} = \frac{k}{f}
  \end{equation*}

  Furthermore, Brown showed that $\mathrm{E}[\Psi]$ and $\mathrm{Var}[\Psi]$ can be calculated directly via a numerical integration:
  \begin{equation*}
    \mathrm{E}[\Psi] = 2k ~~~\text{and}~~~ \mathrm{Var}[\Psi] = 4k + 2\sum_{i<j} \mathrm{Cov}\left( -2\log(P_i), -2\log(P_j) \right)
  \end{equation*}

  Kost and McDermott~\cite{kost_combining_2002} further fit a third-order polynomial to approximate the covariance
  \begin{equation}
    \mathrm{Cov}\left( -2\log(P_i), -2\log(P_j) \right) \approx 3.263 \rho_{ij} + 0.710 \rho_{ij}^2 + 0.027 \rho_{ij}^3
    \label{eqn:covariance-pvalues}
  \end{equation}
  where, $\rho_{ij}$ is the correlation between method $i$ and method $j$

  The final combined p-value~\cite{Poole_Gibbs_Shmulevich_Bernard_Knijnenburg_2016} is then given by:
  \begin{equation}
    \begin{aligned}
        & P_{combined} = 1.0 - \Phi_{2f}\left( \psi / c \right) \\
        \text{where},~ &\psi = -2 \sum_{i=1}^k \log(P_i) ~~~\text{and}~~~ \Phi_{2f} = \mathrm{CDF}\left( \chi^2_{2f} \right)
    \end{aligned}
    \label{eqn:pvalue-combined}
  \end{equation}

  The p-value merging method in \ac{micone} (refer Documentation) uses Equation~\ref{eqn:covariance-pvalues} to estimate the covariance of the pvalues and Equation~\ref{eqn:pvalue-combined} to merge the p-values (obtained from bootstrapping) from the different correlation methods.
  % FIXME: Not sure what our stance on this is?
  Note that we do not use Pearson and Spearman methods in the p-value merging step and these algorithms are only used for demonstration and comparison.
  The combined p-values are used to threshold for significance during the consensus network step.

>>>>

  % QUESTION: Should the CCA section be only included in the response letter?

@@

  % TODO: Create table for this section
  % Should this be in the main manuscript?
  \subsection*{Network metrics}

@@

  \subsection*{p-value merging}

  Fisher~\cite{fisher_224a_1948} proposed that for $w$ independent p-values, each generated by $w$ different methods and denoted by $\bar{P}^i$, the following will hold true for the statistic $\Psi$:
  \begin{equation*}
    \begin{aligned}
      \Psi &= \sum_{i=1}^w -2 \log \left( \bar{P}^i \right) \\
        \Psi &\sim \chi^2_{2w}
    \end{aligned}
  \end{equation*}

  Brown~\cite{brown_400_1975} extended Fisher's method to dependent p-values by using a re-scaled $\chi^2$ distribution:
  \begin{equation*}
    \Psi \sim c \chi^2_{2f}
  \end{equation*}
  where, $f$ is the degrees of freedom and $c$ is the scale factor and are given by:
  \begin{equation*}
    f = \frac{\mathrm{E}[\Psi]^2}{\mathrm{Var}[\Psi]} ~~~\text{and}~~~ c = \frac{\mathrm{Var}[\Psi]}{2\mathrm{E}[\Psi]} = \frac{w}{f}
  \end{equation*}

  Furthermore, Brown showed that $\mathrm{E}[\Psi]$ and $\mathrm{Var}[\Psi]$ can be calculated via a numerical integration:
  \begin{equation*}
    \mathrm{E}[\Psi] = 2w ~~~\text{and}~~~ \mathrm{Var}[\Psi] = 4w + 2\sum_{i<j} \mathrm{Cov}\left( -2\log(\bar{P}^i), -2\log(\bar{P}^j) \right)
  \end{equation*}

  Kost and McDermott~\cite{kost_combining_2002} further fit a third-order polynomial to approximate the covariance
  \begin{equation}
    \mathrm{Cov}\left( -2\log(\bar{P}^i), -2\log(\bar{P}^j) \right) \approx 3.263 \rho_{ij} + 0.710 \rho_{ij}^2 + 0.027 \rho_{ij}^3
    \label{eqn:covariance-pvalues}
  \end{equation}
  where, $\rho_{ij}$ is the correlation between method $i$ and method $j$

  The final combined p-value~\cite{Poole_Gibbs_Shmulevich_Bernard_Knijnenburg_2016} is then given by:
  \begin{equation}
    \begin{aligned}
        & \hat{P}_j = 1.0 - \Phi_{2f}\left( \psi / c \right) \\
        \text{where},~ &\psi = -2 \sum_{i=1}^w \log(\bar{P}^i_j) ~~~\text{and}~~~ \Phi_{2f} = \mathrm{CDF}\left( \chi^2_{2f} \right)
    \end{aligned}
    \label{eqn:pvalue-combined}
  \end{equation}

  The p-value merging and consensus method in \ac{micone} (refer Documentation) uses Equation~\ref{eqn:covariance-pvalues} to estimate the covariance of the pvalues and Equation~\ref{eqn:pvalue-combined} to merge the p-values (obtained from bootstrapping) from the different correlation methods.
  Note that we do not use Pearson and Spearman methods in the p-value merging step and these algorithms are only used for demonstration and comparison.
  The combined p-values are used to threshold for significance during the consensus network step.

  \subsection*{Performance of consensus network on NorTA data}

    \begin{figure}[H]
      \centering
      \includegraphics[width=1.0\linewidth]{figure_s6.pdf}
    \end{figure}
    \begin{figure}[H]
      \centering
        \caption{
        }
      \label{fig:figure_s6}
    \end{figure}
    \FloatBarrier
    \newpage

  \subsection*{Performance of consensus network on seqtime data}

    \begin{figure}[H]
      \centering
      \includegraphics[width=1.0\linewidth]{figure_s7.pdf}
    \end{figure}
    \begin{figure}[H]
      \centering
        \caption{
        }
      \label{fig:figure_s7}
    \end{figure}
    \FloatBarrier
    \newpage
